Composable Concurrency Models

### 2016-11-17 slides/slides.tex
Replaced examples with more crappy drawings

\documentclass{beamer}

\usepackage[USenglish]{babel}
\usepackage[backend=bibtex,citestyle=authortitle]{biblatex}
\usepackage[scale=2]{ccicons}
\usepackage{graphicx}
\usepackage{tikz}

% this actually isn't totally hideous:
\usetheme[numbering=counter, progressbar=frametitle, sectionpage=none]{metropolis}

\graphicspath{{../figures/}}

@@

  \begin{frame}
    \begin{figure}
      \begin{tikzpicture}
        \onslide<1->{
          \node at (0,0) {
            \includegraphics[width=300pt]{browser}
          };
        }

        \onslide<2>{
          \path (-2.25,2.75) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
          \node [text=white] at (-2.25,2.75) {
            Multiple tabs
          };

          \draw (2.25,2.25) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
          \node [text=white] at (2.25,2.25) {
            Suggestions
          };

          \draw (0,0.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
          \node [text=white] at (0,0.5) {
            Page rendering
          };

          \draw (-3,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
          \node [text=white] at (-3,-1.5) {
            Event handling
          };

          \draw (2.5,-1.5) [draw=none, fill=darkgray, opacity=0.75] ellipse (60pt and 25pt);
          \node [text=white] at (2.5,-1.5) {
            Background processes
          }; 
        }

        \onslide<3>{
          \node at (0,-0.62) {
            \includegraphics[width=220pt]{goals}
          };
        }
      \end{tikzpicture}
    \end{figure}
  \end{frame}

@@

    \onslide<1->{
      \centering
      \includegraphics{single-thread-read-write}
    }

    \onslide<2->{
      \centering
      \includegraphics{multiple-threads-read-write}
    }
  \end{frame}

@@

    \begin{itemize}
      \item Linearizability is a guarantee of consistency.
      \item Serializability is a guarantee of isolation.
      \item Together, linearizability and serializability guarantee strict serializability---a history is equivalent to some serial execution order and that order corresponds to the execution order in real time.
    \end{itemize}
  \end{frame}

  \section{Common concurrency models}

  \begin{frame}
    \frametitle{Concurrency models}

    \begin{itemize}
      \item Modern languages usually support several different models.
      \item Common concurrency models can be generalized:~\footcite{Swalens2014}
      \begin{enumerate}
        \item Atomic variables
        \item Software transactional memory (STM)
        \item Communicating threads
        \item Proxies
      \end{enumerate}
    \end{itemize}
  \end{frame}

@@

  \begin{frame}
  \end{frame}

  \subsection{Software transactional memory}

  \begin{frame}
  \end{frame}

@@

  \begin{frame}
  \end{frame}

  \subsection{Proxies}

  \begin{frame}
  \end{frame}

@@

  \subsection{Possible conflicts}

@@

  \section{Conclusion}

  \begin{frame}
    \frametitle{Future work}
  \end{frame}

  \begin{frame}[standout]
    \centering

    \url{github.com/dstelljes/senior-sem}

    \ccbyncsa
  \end{frame}

>>>>

\documentclass[aspectratio=1610]{beamer}

\usepackage{appendixnumberbeamer}
\usepackage[USenglish]{babel}
\usepackage[backend=bibtex,citestyle=authortitle]{biblatex}
\usepackage[scale=2]{ccicons}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{tikz}

% this actually isn't totally hideous:
\usetheme[numbering=counter, progressbar=frametitle, sectionpage=none]{metropolis}
\setbeamertemplate{bibliography item}[text]
\setbeamertemplate{frametitle continuation}{}

@@

  \begin{frame}
    \centering
    \resizebox{0.8\linewidth}{!}{\input{../figures/introduction.tikz}}
  \end{frame}

@@

    \begin{itemize}
      \onslide<1->{
        \item \textbf{Sequential program:} Does the order of operations yield a correct result?
      }

      \onslide<2->{
        \item \textbf{Concurrent program:} Does \emph{every possible} order of operations yield a correct result?
      }
    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Complications}
    \centering

    \textbf{Single thread:}
    \vfill
    \resizebox{0.9\linewidth}{!}{\input{../figures/single-thread-read-write.tikz}}
  \end{frame}

  \begin{frame}
    \frametitle{Complications}
    \centering

    \textbf{Multiple threads:}
    \vfill
    \resizebox{0.9\linewidth}{!}{\input{../figures/multiple-threads-read-write.tikz}}
  \end{frame}

@@

    \begin{itemize}
      \item Linearizability \emph{and} serializability yield strict serializability, which guarantees both consistency and isolation.
    \end{itemize}

    \textbf{Strict serializability~\footcite{Herlihy1990}}

    An ordering of operations is equivalent to some sequential ordering and that ordering corresponds to the order of execution in real time.
  \end{frame}

  \section{Concurrency models}

@@

  \begin{frame}
    \frametitle{Atomic variables}
    \centering

    \includegraphics[width=220pt]{../figures/nonatomic}
  \end{frame}

  \begin{frame}
    \frametitle{Atomic variables}
    \centering

    \includegraphics[width=220pt]{../figures/atomic}
  \end{frame}

  \begin{frame}
    \frametitle{Atomic variables}
    \centering

    \includegraphics[width=220pt]{../figures/lock}
  \end{frame}

  \begin{frame}
    \frametitle{Atomic variables}
    \centering

    \includegraphics[width=220pt]{../figures/multiple-locks}
  \end{frame}

  \subsection{Transactional memory}

  \begin{frame}
    \frametitle{Transactional memory}
    \centering

    \includegraphics[width=220pt]{../figures/stm}
  \end{frame}

  \begin{frame}
    \frametitle{Transactional memory}

    Software transactional memory (STM) is an optimistic approach to working with shared memory:~\footcite{Shavit1995}

    \begin{enumerate}
      \item A thread writes to a shared memory location, keeping track of the transaction in a log.
      \item If there are conflicting changes at the end of the transaction, the transaction is aborted and retried.
      \item If there are no conflicts, the changes are committed and become visible.
    \end{enumerate}
  \end{frame}

@@

  \begin{frame}
    \frametitle{Communicating threads}
    \centering

    \includegraphics[width=220pt]{../figures/agent}
    \vfill
    \textbf{The agent model:} An isolated thread handles all operations on an object.~\footcite{Swalens2014}
  \end{frame}

  \begin{frame}
    \frametitle{Communicating threads}
    \centering

    \includegraphics[width=220pt]{../figures/csp}
    \vfill
    \textbf{Communicating sequential processes:} Independent threads communicate synchronously through predefined channels.~\footcite{Hoare1978}
  \end{frame}

  \begin{frame}
    \frametitle{Communicating threads}
    \centering

    \includegraphics[width=220pt]{../figures/actor}
    \vfill
    \textbf{The actor model:} Independent threads send messages to known addresses.~\footcite{Agha1986}
  \end{frame}

@@

  \begin{frame}
    \frametitle{Composability criteria}

    \textbf{How do we know that models are composable?}

    \begin{itemize}
      \item Safety: ``Nothing bad will happen!'' (The output of a program or algorithm will not be incorrect.)
      \item Liveness: ``Something will eventually happen!'' (The program or algorithm will terminate.)
    \end{itemize}

    Two models are composable if using them within each other doesn't compromise safety or liveness.~\footcite{Swalens2014}
  \end{frame}

  \begin{frame}
    \frametitle{Composability criteria}
    \centering

    \includegraphics[width=360pt]{../figures/clojure-survey}
  \end{frame}

  \begin{frame}
    \frametitle{Possible conflicts}

    \begin{itemize}
      \onslide<1->{
        \item A model reexecutes code containing another model that performs an irrevocable action.
      }
      \vfill
      \onslide<2->{
        \item A model reexecutes code that causes the reexecution to continually happen.
      }
      \vfill
      \onslide<3->{
        \item A model that supports blocking operations is used within a model that does not support blocking operations.
      }
      \vfill
      \onslide<4->{
        \item A model does not guarantee safety or liveness by design, making composability impossible.
      }
    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Conclusions}

    \begin{itemize}
      \onslide<1->{
        \item The language may ensure that some bad interactions don't happen.
      }
      \vfill
      \onslide<2->{
        \item Sometimes, it's best to just use a different concurrency model.
      }
      \vfill
      \onslide<3->{
        \item Future work would identify common ``building blocks'' (thread creation, message passing, etc.) that could be composed to express existing concurrency models and combinations of models.
      }
    \end{itemize}
  \end{frame}

@@

  \begin{frame}
    \frametitle{Proposed abstractions}

    \textbf{What characteristics should a unified abstraction have?}~\footcite{Marr2012}

    \begin{itemize}
      \onslide<1->{
        \item Ownership: All objects are owned by something, so ownership should be supported in a way that supports regulated access.
      }
      \vfill
      \onslide<2->{
        \item Managed mutation and execution: The owner of an object should determine how it be read and modified.
      }
      \vfill
      \onslide<3->{
        \item Leveled access: Threads should either have model-level access or abstraction-level access.
      }
      \vfill
      \onslide<4->{
        \item Enforceability: The model's semantics should be enforced regardless of where the model is used.
      }
    \end{itemize}
  \end{frame}

  \begin{frame}
    \frametitle{Ownership-based meta-object protocol}

    \centering

    \includegraphics[width=250pt]{../figures/obmop-simple.png}
  \end{frame}

  \begin{frame}
    \frametitle{Conclusions}
  \end{frame}

  \begin{frame}[standout]
    \centering
    Thanks to Elena Machkasova, K.K. Lamberty, and Matthew Justin for their guidance and suggestions.
    \vfill
    \href{https://github.com/dstelljes/senior-sem}{github.com/dstelljes/senior-sem}
    \vfill
    \ccbyncsa{}
  \end{frame}

  \appendix{}

  \begin{frame}[allowframebreaks]
    \frametitle{References}

    \printbibliography{}
  \end{frame}
